---
title: "Maine Healthy Beaches: Data Exploration"
author: "Jeffrey D Walker, PhD"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document: 
    css: styles.css
    number_sections: yes
    toc: yes
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(ggplot2)
theme_set(theme_bw())
library(gridExtra)
library(knitr)

opts_chunk$set(echo=FALSE, message=FALSE)
```

```{r load-data, cache=TRUE, results='hide'}
stn <- readRDS("../data/rdata/stn.rda")
ent <- readRDS("../data/rdata/wq-resamples.rda")
ent <- mutate(ent,
              MONTH=month(SAMPLE_DATE),
              GT_70=ENTEROCOCCI > 70,
              GT_104=ENTEROCOCCI > 104,
              CLASS=ordered(GT_70+GT_104, levels=c(0, 1, 2),
                            labels=c('<=70', '70-104', '>104')),
              SAMPLE_TYPE=ifelse(RESAMPLE, "Resample",
                                 ifelse(OTHER, "Irregular", "Routine")),
              SAMPLE_TYPE=ordered(SAMPLE_TYPE, levels=c("Routine", "Resample", "Irregular")))
ent <- left_join(ent,
                 select(stn, TOWN, BEACH_NAME=SITE_NAME, SAMPLE_POINT_NAME) %>% distinct,
                 by="SAMPLE_POINT_NAME")

scale_class_fill <- scale_fill_manual('', values=c('<=70'='grey50', '70-104'='deepskyblue', '>104'='orangered'))
scale_class_color <- scale_color_manual('', values=c('<=70'='grey50', '70-104'='deepskyblue', '>104'='orangered'))

ent_limits <- data.frame(ENTEROCOCCI=c(70, 104), LABEL=ordered(c('70', '104')))
scale_limits_color <- scale_color_manual('', values=c('70'='deepskyblue', '104'='orangered'))
```


```{r plot-functions}
log_breaks <- function(x, y) {
  # log breaks
  as.vector(outer(x, y, '*'))
}
log_labels <- function(x, y) {
  # labels for log scales with gaps
  x_na <- seq(1, 9)
  x_na[which(!(x_na %in% x))] <- NA
  x <- log_breaks(x_na, y)
  x <- as.character(x)
  x <- ifelse(is.na(x), "", x)
  x
}
log_x <- scale_x_log10(breaks=log_breaks(seq(1, 9), 10^seq(-3, 5)),
                       labels=log_labels(c(1, 5), 10^seq(-3, 5)))
log_y <- scale_y_log10(breaks=log_breaks(seq(1, 9), 10^seq(-3, 5)),
                       labels=log_labels(c(1, 5), 10^seq(-3, 5)))
```


# Distribution of Enterococcus Concentrations

## All Samples

This section summarizes the distribution of all enterococcus samples. It only includes routine samples and thus excludes all resamples.

This figure shows the histogram of all routine samples on a log-scale. The vertical lines show the 70 and 104 limits. The distribution is uni-modal, the large change at 10 MPN/100mL is due to the detection limits.

```{r plot-histogram}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  ggplot() +
  geom_histogram(aes(ENTEROCOCCI), binwidth=0.3) +
  geom_vline(aes(xintercept=ENTEROCOCCI, color=LABEL),
             data=ent_limits) +
  scale_limits_color +
  scale_y_continuous(labels=scales::comma) +
  log_x +
  labs(x="ENT", y="# Samples",
       title="Distribution of All Samples")
```

This figure shows the same data, but on a linear scale and only including samples with ENT < 200 to focus on the lower range. Again, the majority of routine samples are at or below the detection limit of 10 MPN/100mL.

```{r plot-histogram-200}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  filter(ENTEROCOCCI < 200) %>%
  ggplot() +
  geom_histogram(aes(ENTEROCOCCI), binwidth=10) +
  geom_vline(aes(xintercept=ENTEROCOCCI, color=LABEL), data=ent_limits) +
  scale_limits_color +
  scale_y_continuous(labels=scales::comma) +
  labs(x="ENT (MPN/100mL)", y="# Samples",
       title="Distribution of All Samples with ENT < 200")
```

**Question**: What fraction of all routine samples are between 70 and 104?

**Answer**: `r scales::percent(prop.table(table(filter(ent, SAMPLE_TYPE == "Routine")$CLASS))['70-104'])`

This table shows the fraction of routine samples <70, between 70-104, and >104.

```{r prop-class}
ent %>% 
  filter(SAMPLE_TYPE == "Routine") %>%
  select(CLASS) %>% 
  table() %>%
  prop.table() %>%
  as.data.frame %>%
  mutate(Freq=scales::percent(Freq)) %>%
  kable(col.names=c('Range', 'Percent'), digits=3)
```


**Question**: If a routine sample is greater than 70 MPN/100mL, what is the probability it is also > 104?

**Answer**: `r scales::percent(prop.table(table(filter(ent, GT_70, SAMPLE_TYPE == "Routine")$CLASS))['>104'])`

This table shows the fraction of samples between 70-104 and >104 based only on samples with concentrations <70. This shows that out of all samples greater than 70 MPN/100mL, about 70% are also above 104, and 30% are between 70 and 104.

```{r prop-class-gt70}
ent %>% 
  filter(SAMPLE_TYPE == "Routine") %>%
  filter(GT_70) %>%
  droplevels %>% 
  select(CLASS) %>% 
  table() %>%
  prop.table() %>%
  as.data.frame %>%
  mutate(Freq=scales::percent(Freq)) %>%
  kable(col.names=c('Range', 'Percent'), digits=3)
```

## By Beach

In this section, the data are grouped by beach and thus may include more than one sampling site.

This figure shows the total number of routine samples in each category (<70, 70-104, >104).

```{r plot-count-beach, fig.width=10, fig.height=10}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(BEACH_NAME) %>%
  mutate(N_BEACH = n()) %>%
  ungroup %>%
  arrange(N_BEACH) %>%
  mutate(BEACH_NAME=ordered(BEACH_NAME, levels=unique(BEACH_NAME))) %>%
  group_by(BEACH_NAME, CLASS) %>%
  summarise(N=n()) %>%
  ungroup %>%
  spread(CLASS, N, fill=0) %>%
  gather(CLASS, N, -BEACH_NAME) %>%
  mutate(CLASS=ordered(CLASS, levels=rev(levels(ent$CLASS)))) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  arrange(desc(CLASS)) %>%
  ggplot(aes(BEACH_NAME, N, fill=CLASS)) +
  geom_bar(stat='identity', position='stack') +
  scale_class_fill +
  labs(y="Number of Samples", x="Beach",
       title="Number of Samples by Beach and Category") +
  coord_flip() +
  theme(axis.text.y=element_text(size=6))
```

This figure shows the same data, but instead of the number of routine samples, it shows the fraction of routine samples in each category by beach ordered by the fraction >104 (dirtiest beaches at the top, cleanest at the bottom). 

```{r plot-count-beach-pct, fig.width=10, fig.height=10}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(BEACH_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  select(-N) %>%
  spread(CLASS, FRAC, fill=0) %>%
  arrange(`>104`) %>%
  mutate(BEACH_NAME=ordered(BEACH_NAME, levels=unique(BEACH_NAME))) %>%
  gather(CLASS, FRAC, -BEACH_NAME) %>%
  mutate(CLASS=ordered(CLASS, levels=rev(levels(ent$CLASS)))) %>%
  arrange(desc(CLASS)) %>%
  ggplot(aes(BEACH_NAME, FRAC, fill=CLASS)) +
  geom_bar(stat='identity', position='stack') +
  scale_class_fill +
  scale_y_continuous(labels=scales::percent) +
  labs(y="% of Samples", x="Beach",
       title="Fraction of Samples by Beach and Category") +
  coord_flip() +
  theme(axis.text.y=element_text(size=6))
```

This figure shows only the fraction of routine samples between 70-104 (the blue bars of the previous figure) and ordered from highest to lowest. 

```{r plot-beach-count-pct-70-104, fig.width=10, fig.height=10}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(BEACH_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  select(-N) %>%
  spread(CLASS, FRAC, fill=0) %>%
  arrange(`70-104`) %>%
  mutate(BEACH_NAME=ordered(BEACH_NAME, levels=unique(BEACH_NAME))) %>%
  gather(CLASS, FRAC, -BEACH_NAME) %>%
  mutate(CLASS=ordered(CLASS, levels=rev(levels(ent$CLASS)))) %>%
  arrange(desc(CLASS)) %>%
  filter(CLASS=="70-104") %>%
  ggplot(aes(BEACH_NAME, FRAC, fill=CLASS)) +
  geom_bar(stat='identity', position='stack') +
  scale_class_fill +
  scale_y_continuous(labels=scales::percent, breaks=seq(0, 1, by=0.01)) +
  labs(y="% of Samples", x="Beach",
       title="Fraction of Samples between 70-104 by Beach") +
  coord_flip() +
  theme(axis.text.y=element_text(size=6))
```

This figure shows the cumulative distribution of % routine samples for each category across beaches. It shows, for example, that 50% of the beaches have 90% or less of samples <70 (dotted gray line), and that 90% of beaches have 7% or less of samples between 70-104 (dotted blue line).

```{r plot-beaches-cdf, fig.width=8, fig.height=6}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(BEACH_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  arrange(CLASS, FRAC) %>%
  group_by(CLASS) %>%
  mutate(INDEX=row_number()/n()) %>%
  ggplot(aes(INDEX, FRAC, color=CLASS)) +
  geom_point() +
  geom_line() +
  geom_segment(x=0, xend=0.5, y=0.8965, yend=0.8965, color='grey50', linetype='dotted') +
  geom_segment(x=0.5, xend=0.5, y=0, yend=0.8965, color='grey50', linetype='dotted') +
  geom_segment(x=0, xend=0.9, y=0.068085106, yend=0.068085106, color='deepskyblue', linetype='dotted') +
  geom_segment(x=0.9, xend=0.9, y=0, yend=0.068085106, color='deepskyblue', linetype='dotted') +
  scale_class_color +
  scale_x_continuous(labels=scales::percent, breaks=seq(0, 1, 0.1)) +
  scale_y_continuous(labels=scales::percent, breaks=seq(0, 1, 0.1)) +
  labs(x="% of Beaches", y="% of Samples",
       title="Cumulative Distribution of Fraction Samples by Beach and Category")
```

This figure shows the same data, but separates each category into separate panels with different y-axis scales.

```{r plot-beaches-cdf-facet, fig.width=10, fig.height=4}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(BEACH_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  # filter(CLASS=='70-104') %>%
  arrange(CLASS, FRAC) %>%
  group_by(CLASS) %>%
  mutate(INDEX=row_number()/n()) %>%
  ggplot(aes(INDEX, FRAC, color=CLASS)) +
  geom_point() +
  geom_line() +
  scale_class_color +
  scale_x_continuous(labels=scales::percent, breaks=seq(0, 1, 0.1)) +
  scale_y_continuous(labels=scales::percent) +
  labs(x="% of Beaches", y="% of Samples",
       title="Cumulative Distribution of Fraction Samples by Beach and Category with Panels") +
  facet_wrap(~CLASS, scales='free_y') +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
```


## By Site

In this section, the data are grouped by site instead of beach.

This figure shows the total number of samples in each category (<70, 70-104, >104), ordered by the total number of samples.

```{r plot-count-site, fig.width=10, fig.height=12}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(SAMPLE_POINT_NAME, CLASS) %>%
  summarise(N=n()) %>%
  ungroup %>%
  spread(CLASS, N, fill=0) %>%
  gather(CLASS, N, -SAMPLE_POINT_NAME) %>%
  mutate(CLASS=ordered(CLASS, levels=rev(levels(ent$CLASS)))) %>%
  mutate(FRAC=N/sum(N)) %>%
  group_by(SAMPLE_POINT_NAME) %>%
  mutate(N_SITE=sum(N)) %>%
  ungroup %>%
  arrange(N_SITE) %>%
  mutate(SAMPLE_POINT_NAME=ordered(SAMPLE_POINT_NAME, levels=unique(SAMPLE_POINT_NAME))) %>%
  arrange(desc(CLASS)) %>%
  ggplot(aes(SAMPLE_POINT_NAME, N, fill=CLASS)) +
  geom_bar(stat='identity', position='stack') +
  scale_class_fill +
  labs(y="Number of Samples", x="Beach",
       title="Number of Samples by Site and Category") +
  coord_flip() +
  theme(axis.text.y=element_text(size=6))
```

This figure shows the same data, but instead of the number of samples, it shows the fraction of samples in each category by beach ordered by the fraction >104 (dirtiest beaches at the top, cleanest at the bottom). 

```{r plot-count-site-pct, fig.width=10, fig.height=10}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(SAMPLE_POINT_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  group_by(SAMPLE_POINT_NAME) %>%
  mutate(N_SITE=sum(N)) %>%
  ungroup %>%
  select(-N, -N_SITE) %>%
  spread(CLASS, FRAC, fill=0) %>%
  arrange(`>104`) %>%
  mutate(SAMPLE_POINT_NAME=ordered(SAMPLE_POINT_NAME, levels=unique(SAMPLE_POINT_NAME))) %>%
  gather(CLASS, FRAC, -SAMPLE_POINT_NAME) %>%
  mutate(CLASS=ordered(CLASS, levels=rev(levels(ent$CLASS)))) %>%
  arrange(desc(CLASS)) %>%
  ggplot(aes(SAMPLE_POINT_NAME, FRAC, fill=CLASS)) +
  geom_bar(stat='identity', position='stack') +
  scale_class_fill +
  scale_y_continuous(labels=scales::percent) +
  labs(y="% of Samples", x="Site",
       title="Fraction of Samples by Beach and Category") +
  coord_flip() +
  theme(axis.text.y=element_text(size=6))
```

This figure shows only the fraction of total samples between 70-104 (the blue bars of the previous figure) and ordered from highest to lowest. 

```{r plot-site-count-pct-70-104, fig.width=10, fig.height=10}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(SAMPLE_POINT_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  select(-N) %>%
  spread(CLASS, FRAC, fill=0) %>%
  arrange(`70-104`) %>%
  mutate(SAMPLE_POINT_NAME=ordered(SAMPLE_POINT_NAME, levels=unique(SAMPLE_POINT_NAME))) %>%
  gather(CLASS, FRAC, -SAMPLE_POINT_NAME) %>%
  mutate(CLASS=ordered(CLASS, levels=rev(levels(ent$CLASS)))) %>%
  arrange(desc(CLASS)) %>%
  filter(CLASS=="70-104") %>%
  ggplot(aes(SAMPLE_POINT_NAME, FRAC, fill=CLASS)) +
  geom_bar(stat='identity', position='stack') +
  scale_class_fill +
  scale_y_continuous(labels=scales::percent, breaks=seq(0, 1, by=0.01)) +
  labs(y="% of Samples", x="Site",
       title="Fraction of Samples between 70-104 by Site and Category") +
  coord_flip() +
  theme(axis.text.y=element_text(size=6))
```

This figure shows the cumulative distribution of % samples for each category across beaches. It shows, for example, that 50% of the beaches have 90% or less of samples <70 (dotted gray line), and that 90% of beaches have 7% or less of samples between 70-104 (dotted blue line).

```{r plot-sites-cdf, fig.width=8, fig.height=6}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(SAMPLE_POINT_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  arrange(CLASS, FRAC) %>%
  group_by(CLASS) %>%
  mutate(INDEX=row_number()/n()) %>%
  ggplot(aes(INDEX, FRAC, color=CLASS)) +
  geom_point() +
  geom_line() +
  geom_segment(x=0, xend=0.5, y=0.9014, yend=0.9014, color='grey50', linetype='dotted') +
  geom_segment(x=0.5, xend=0.5, y=0, yend=0.9014, color='grey50', linetype='dotted') +
  geom_segment(x=0, xend=0.9, y=0.0714, yend=0.0714, color='deepskyblue', linetype='dotted') +
  geom_segment(x=0.9, xend=0.9, y=0, yend=0.0714, color='deepskyblue', linetype='dotted') +
  scale_class_color +
  scale_x_continuous(labels=scales::percent, breaks=seq(0, 1, 0.1)) +
  scale_y_continuous(labels=scales::percent, breaks=seq(0, 1, 0.1)) +
  labs(x="% of Sites", y="% of Samples",
       title="Cumulative Distribution of Fraction Samples by Site and Category")
```

This figure shows the same data, but separates each category into separate panels with different y-axis scales.

```{r plot-sites-cdf-facet, fig.width=10, fig.height=4}
ent %>%
  filter(SAMPLE_TYPE == "Routine") %>%
  group_by(SAMPLE_POINT_NAME, CLASS) %>%
  summarise(N=n()) %>%
  mutate(FRAC=N/sum(N)) %>%
  ungroup %>%
  arrange(CLASS, FRAC) %>%
  group_by(CLASS) %>%
  mutate(INDEX=row_number()/n()) %>%
  ggplot(aes(INDEX, FRAC, color=CLASS)) +
  geom_point() +
  geom_line() +
  scale_class_color +
  scale_x_continuous(labels=scales::percent, breaks=seq(0, 1, 0.1)) +
  scale_y_continuous(labels=scales::percent) +
  labs(x="% of Sites", y="% of Samples",
       title="Cumulative Distribution of Fraction Samples by Site and Category with Panels") +
  facet_wrap(~CLASS, scales='free_y') +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
```


# Sampling Frequency and Resampling

This section looks at how often resamples come back above or below the current limit of 104. This figure plots the each resample concentration against the concentration of the previous (routine) sample.

```{r}
ent %>%
  arrange(SAMPLE_POINT_NAME, SAMPLE_DATE) %>%
  group_by(SAMPLE_POINT_NAME, YEAR) %>%
  mutate(PREV_ENTEROCOCCI=lag(ENTEROCOCCI),
         PREV_CLASS=lag(CLASS),
         DT=as.numeric(difftime(floor_date(SAMPLE_DATE, "day"),
                                lag(floor_date(SAMPLE_DATE, "day")),
                                units="days"))) %>%
  # select(SAMPLE_POINT_NAME, SAMPLE_DATE, ENTEROCOCCI, PREV_ENTEROCOCCI,
  #        CLASS, PREV_CLASS, DT) %>%
  filter(RESAMPLE, PREV_ENTEROCOCCI>104, DT<=3) %>%
  ggplot(aes(PREV_ENTEROCOCCI, ENTEROCOCCI, color=CLASS)) +
  geom_point() +
  geom_hline(aes(yintercept=ENTEROCOCCI),
             data=ent_limits, linetype=2) +
  scale_class_color +
  log_x + 
  log_y +
  labs(x="Previous Routine Enterococci (MPN/100mL)",
       y="Resample Enterococci (MPN/100mL)",
       title="Comparison of Resample Concentration to Previous Sample")
```

**Question**: What percent of all resamples have concentrations exceeding 104?

```{r}
ent_resample <- ent %>%
  arrange(SAMPLE_POINT_NAME, SAMPLE_DATE) %>%
  group_by(SAMPLE_POINT_NAME, YEAR) %>%
  mutate(PREV_ENTEROCOCCI=lag(ENTEROCOCCI),
         PREV_CLASS=lag(CLASS),
         DT=as.numeric(difftime(floor_date(SAMPLE_DATE, "day"),
                                lag(floor_date(SAMPLE_DATE, "day")),
                                units="days"))) %>%
  filter(RESAMPLE, !(lag(RESAMPLE)), PREV_ENTEROCOCCI>104, DT<=3) %>%
  group_by(SAMPLE_POINT_NAME) %>%
  summarise(N_TOTAL=n(),
            N_EXCEED=sum(ENTEROCOCCI > 104)) %>%
  summarise(N_TOTAL=sum(N_TOTAL),
            N_EXCEED=sum(N_EXCEED))
```

**Answer**: There were `r ent_resample$N_EXCEED` resamples that exceed 104 MPN/100mL out of `r ent_resample$N_TOTAL` total resamples, and thus `r scales::percent(ent_resample$N_EXCEED/ent_resample$N_TOTAL)` of all resamples continue to exceed 104 MPN/100mL.

This figure shows the percent of all resamples that come back above the current limit of 104 by site. This only includes initial resamples (i.e. resamples that follow a routine sample, but not resamples that are the second or third resample in a row). This also only includes sites that have at least 10 resamples total over the entire period (to exclude sites with only a couple resamples). This shows that at LIN-3, more than 40% of the resamples following a routine sample continued to exceed the 104 MPN/100mL.

```{r}
ent %>%
  arrange(SAMPLE_POINT_NAME, SAMPLE_DATE) %>%
  group_by(SAMPLE_POINT_NAME, YEAR) %>%
  mutate(PREV_ENTEROCOCCI=lag(ENTEROCOCCI),
         PREV_CLASS=lag(CLASS),
         DT=as.numeric(difftime(floor_date(SAMPLE_DATE, "day"),
                                lag(floor_date(SAMPLE_DATE, "day")),
                                units="days"))) %>%
  # select(SAMPLE_POINT_NAME, SAMPLE_DATE, ENTEROCOCCI, PREV_ENTEROCOCCI,
  #        CLASS, PREV_CLASS, DT) %>%
  filter(RESAMPLE, !(lag(RESAMPLE)), PREV_ENTEROCOCCI>104, DT<=3) %>%
  group_by(SAMPLE_POINT_NAME) %>%
  summarise(N_TOTAL=n(),
            N_EXCEED=sum(ENTEROCOCCI > 104)) %>%
  mutate(FRAC=N_EXCEED/N_TOTAL) %>%
  filter(N_TOTAL > 10) %>%
  arrange(desc(FRAC)) %>%
  mutate(SAMPLE_POINT_NAME=ordered(SAMPLE_POINT_NAME, levels=rev(SAMPLE_POINT_NAME))) %>%
  ggplot(aes(SAMPLE_POINT_NAME, FRAC)) +
  geom_bar(aes(fill=">104"), stat="identity") +
  scale_y_continuous(labels=scales::percent, breaks=seq(0, 1, by=0.1)) +
  scale_fill_manual("", values="orangered") +
  labs(y="% of Resamples Exceeding 104", x="Site",
       title="Fraction of Resamples Exceeding 104 by Site") +
  coord_flip() +
  theme(axis.text.y=element_text(size=6))
```

